\pdfbookmark[0]{Acknowledgment}{Acknowledgment}
\chapter*{Acknowledgment}
\label{chp:acknowledgment}
\vspace*{-10mm}

%\lipsum
I would like to express my gratitude to all who have accompanied me throughout the years of research as a PhD student, in particular to:

My supervisor Prof. Dr.-Ing. Martin Gaedke for introducing me to the challenging but exciting world of research, for his inspiring vision and ideas, and for giving me the opportunity to learn and grow under his guidance in a supportive and motivating environment that he created and to collaborate with many international researchers. 

To all my colleagues and dear friends with whom I had the honor to collaborate on various research projects, who have widened my research interests and from whom I have learned so many things, I owe so much. 
Here, I would like to especially mention Maxim Bakaev, Mahda Noura, Stefan Wild, Alexey Tschudnowsky, Julian Grigera, José Matías Rivero, Bahareh Zarei, and Maximilian Speicher.

I am very grateful to all students and colleagues who contributed to my research presented in this thesis: Felix Förster, Valentin Siegert, Tobias Lang, Thomas Weber, Frank Siegel, Anna Scholz, Philipp Oehme, Markus Keller, Stefan Staude, and Hao Li, and to Ralph Sontag, who has always encouraged me and spent countless hours helping me to typeset this thesis. I cordially thank my friend Mahsa Sodeiri for helping me design the book cover for the print version of this thesis.

I thank all present and former colleagues at VSR -- André Langer, Maik Benndorf, Michael Krug, Fabian Wiedemann, Jörg Anders, Markus Ast, Christoph Brandt, Dang Vu Nguyen Hai, Felix Kettner, and Verena Traubinger -- for inspiring and fruitful discussions, continuous knowledge sharing, and a friendly and welcoming atmosphere that has lead me through the harder times in my PhD student life. Prof. Dr.-Ing. Martin Gaedke did an excellent job in putting the VSR team together.

Thank you also to my three reviewers for their encouraging feedback and support to successfully write, defend, and publish this thesis, as well as to the members of the PhD  commission, Prof. Dr.-Ing. Janet Siegmund, Dr. Andreas Müller, and Dr. Frank Seifert.

Also, I like to thank all student assistants and students advised by me for enabling me to focus on my research and teaching me how to communicate ideas and organize research activities.

Finally, I am very grateful to my mother for her ongoing encouragement, appreciation, understanding, love and care, and to Michalina Kunecka, Sven R. Kunze, Christian Jager, Bastian Weiß, Gabriela Bosetti, Radka Nacheva and many more of my friends whom I have not already mentioned above, who all supported me, raised me up and kept close contact in spite of the sometimes limited availability resulting from the PhD challenge.

\cleardoublepage