\hypertarget{sec:introduction}{%
\chapter{Introduction}\label{sec:introduction}}
In the era of overwhelming online data, enabling non-technical users to
make more informed decisions has become a critical priority. A factor
determinant of the decision preciseness and accuracy is the extent to
which the decision support tool offers domain specificity \autocite{Zhu2022}. However, the rapid technological advancements and
ever-changing domain requirements, coupled with the vast accessible
online data, have made situational decision-making more challenging and
prone to inaccuracies. Low expressive power, lack of customizability,
and unreasonable usability of current situational decision-making tools
are main contributing problems. Addressing these challenges is essential
to fully resolving the central problem. This section describes the
current situation and motivations for developing situational
decision-making tools. It also outlines the central problem and the
contributing subproblems that form the basis for this thesis. 


\vspace{-15pt}
\hypertarget{sec:situation}{%
\section{Situation}\label{sec:situation}}
\vspace{15pt}

This section provides an overview of current situation in which the
challenge of domain specificity in situational decision-making occurs
for domain experts. Situational decision-making concerned with making
effective and rapid decisions in a specific context and domain to
address diverse requirements \autocite{Hagele2021}. Based
on this definition, situational applications are characterized by having
short lifespan and being contextually customizable to address immediate
and non-permanent requirements of domain-experts \autocite{Pahlke2010}. Reflecting the ever-changing requirements in enterprise
management ecosystems, situational applications are becoming
increasingly important. Given the peculiar characteristics of these
applications, traditional software development processes are neither
sufficient nor reasonable for their design and creation \autocite{Casati2012}.

Promising advances in communication and information technology during
the last decades result in overwhelming amount of data available online
\autocite{Noran2018}. To contribute to this situation, increasing
number of service providers publish functionalities and data on public
registries as web services. According to \autocite{Zarei2021}, as end
of 2021, every year 2000 new web services was registered on
ProgrammableWeb. Such advancements are influencing decision making
paradigm in various sectors by promising adequate knowledge to aid
domain experts making situational decisions more informatively \autocite{Noran2018}.

Domain experts in the context of this work are non-technical end-users
who possess adequate domain knowledge and capable of performing
domain-related operations. Involving domain experts in development
process of situational applications, brought advantages such as
improving end tool's provision and accelerating implementation and
development cycles. Moreover, domain experts can propose levels of
creativity which cannot be introduced by developers due to their lack of
domain knowledge \autocite{Ponce2022}. To fully realize the domain experts potential, two research fields,
\gls{hci} and \gls{eud}, propose
context-aware solutions that bridge the gap between developers and
domain experts \autocite{Wei2020}. \gls{eud} methods targets end user
as the solution designer and developer. According to \autocite{Lieberman2006}, \gls{eud} can be defined as ``Set of methods, techniques and
tools to enable non-technical end-users to create and modify software
artifacts.'' Based on this definition, \gls{eud} proposes unique competitive
advantage for domain-experts by enabling them to address their
situational needs in dynamic environments \autocite{HenryLiebermanFabioPaterno2004}.

A crucial focus point for domain experts is the tool's domain
specificity, which is a well-known mean to ensure the application
adoption in a domain-specific context \autocite{Ardito2018}. A
domain-specific tool involves domain experts only in activities,
operations and concepts that are restricted to their area of expertise
\autocite{Soi2014a}. In the terms of domain specificity, the \gls{eud}
approaches face challenges in different development phases to target
multiple levels of customizability for the envisioned situational tool
\autocite{Ponce2022}.

Most promising \gls{eud} solutions within this field are in the form of Web
Mashups (also known as mashups or service-mashup). Web Mashup proposes
easy-to-use and light weight composition of available services on the
web (such as \gls{api}s) allowing end-users to access multiple data sources
and service, in a single location \autocite{Santos2019}. A number
of mashup tools have been developed by leading companies, such as IBM
Mashup Center. Given the web mashup's popularity and the characteristics
of situational applications, web mashups can offer a viable solution for
situational applications \autocite{Cao2017}.


\vspace{-15pt}
\hypertarget{motivation}{%
\section{Motivation}\label{motivation}}
\vspace{15pt}

Despite technological advancements in data generation and functionalities access (via \gls{api}s, web services, etc.), the disparity between information provided by online services and the knowledge necessary for situational decision-making is widening \autocite{Noran2018}.A large number of systems designed to enable end users making more informed decisions, with Business Intelligence (BI) holding a central significance. The term BI refers to a group of software applications, platforms, technologies, and processes for assisting end-users in decision-making. However, the provided solutions are not tailored to the unique requirements of situational applications therefore, cannot fully satisfy the customizability and domain specificity required by domain experts \autocite{Pahlke2010} \autocite{Ardito2018}.
On the other hand, developing situational applications requires a specific technical expertise that domain experts do not possess. Technical skills in this context range from understanding web technologies, such as\gls{api} integration and scripting languages, to domain modeling and tool customization \autocite{Ponce2022}. From the programmers' perspective, developing applications for transient requirements from scratch is deemed unprofitable in terms of time and effort \autocite{Aghaee2014}. Although numerous solutions in the field of \gls{eud} have been proposed in recent years to address development challenges for non-technical users, only a few primarily focus on domain specificity. According to a survey conducted in 2019, only 33\% of \gls{eud} solutions are domain-specific, with two-thirds of solutions concentrating on providing general applications \autocite{Santos2019}. Bearing in mind that adopting such generic applications to be domain-specific is a non-trivial and error prone task \autocite{Soi2014a}. Therefore, a suitable approach is required to empower domain experts in developing situational applications in compliance with their domain of expertise.


\vspace{-15pt}
\hypertarget{sec:problem}{%
\section{Problem Statement}\label{sec:problem}}
\vspace{15pt}

Concerning the situation described in the previous chapter, the identified central problem in this thesis is “High complexity and technical barriers in developing situational decision-support systems within a well-defined domain”. These systems are software solutions providing functionalities required for domain-specific decision-making. The immediate effect of the problem is manifested as domain experts’ failure in making informed decisions suitable for domain’s peculiar characteristics. 
By thoroughly investigating the central problem, three contributing subproblems are identified.

\textbf{Limited usability:} Usability \autocite[cf.][def. 3.4451]{ISO/IEEE24765Vocabulary} is considered as an important
quality factor in terms of web applications which mainly targets the
human-computer interactions \autocite{Ponce2022}. The notion of ``easy to use''
determines a tool's usability which is principally related to efficiency
and effectiveness \autocite{Zhao2019}. Achieving proper level of
usability for domain experts is mostly hindered by the overwhelming
complexity of emerging web technologies and amount of data available
online. This issue manifests itself mainly in complex scenarios of
situational applications \autocite{Ponce2022} \autocite{Tang2021}. EUD
proposes user-centric, lightweight approaches to foster usability for
domain experts. Despite number of various solutions in this domain, the
current EUD tools are still too difficult to be used for situational
decision making \autocite{Zhao2019}. The immediate effect of this
subproblem manifests itself in uninformative and poor situational
decisions made by domain experts.

\textbf{Poor expressive power:} The definition of expressiveness varies
according to the application domain. Generally, a tool's
expressive power is determined by various solution classes that can be
generated by domain experts enabling them to go beyond trivial
application logic \autocite{Soi2013}. In the context of this work a tool is
considered expressive if it can comprehend domain expert intentions and
reflect them in the final solutions accordingly. The gap between
situational requirements and final solution occurs because tools do not
speak the domain expert's language. To contribute to this issue, most
existing EUD tools sacrifices their expressivity to achieve higher
usability \autocite{Ponce2022}. Therefore, restricted expressiveness results in
solutions with trivial logic incapable of assisting domain experts with
their complex situational requirements.

\textbf{Limited domain specificity}. Given situational application's peculiar characteristics, domain
specificity plays a crucial role in guaranteeing the effectiveness of
the final decision within a specific domain \autocite{Noran2018}.
Domain specificity influences constructing situational applications by
integrating domain knowledge as an essential element. The domain
knowledge helps in assessing the solution accuracy, dependability, and
various other quality metrics \autocite{Bryant2010}. Despite the
mentioned benefits, the utilization of domain-specific situational tools
faces impediments, and few numbers of research are dedicated to this
issue. Key challenges include the developer's limited
domain knowledge and the lack of technical skills among domain experts,
both of which are critical for integrating knowledge graphs and
fostering domain-specific applications \autocite{Santos2019}
\autocite{Casati2012}. Other contributing factor includes the
absence of standardized domain models to capture the essential
domain-related knowledge for efficient decision making \autocite{Ponce2022}.
This problem results in having unfamiliar situational applications with
high learning curve. Moreover, these tools lack the accuracy required
for domain-specific tasks. The detailed causes and effects to the
identified problems are presented as a Problem Tree (cf. Appendix A). A
problem tree in the context of Logical Framework Approach (LFA),
presents the foundation for systematic and scientific procedure \autocite{EUIntegrationOffice.2011}.

\vspace{-15pt}
\hypertarget{sec:objectives}{%
\section{Research Objectives}\label{sec:objectives}}
\vspace{15pt}

To derive solutions for the above-mentioned central problem, a series of research objectives is designed. The primary goal of this thesis is assisting domain experts in making situational decisions within a well-defined domain by developing a time and effort efficient composition platform. The solution should have a dedicated emphasize on usability, domain specificity and expressiveness measures. To achieve this goal, three objectives are addressed in this thesis. Each objective contributes to the overall solution by targeting usability, in the sense of providing an easy-to-use composition platform; expressiveness, by reducing the gap between domain expert’s intentions and end result; and domain specificity, with supporting domain models and domain-related customizability.

\begin{researchobjective}{ro:1}
Enable domain experts to develop an easy-to-use situational application and minimize reliance on software developers. 
\end{researchobjective}

\begin{researchobjective}{ro:2}
Provide a natural communication channel for domain experts to express their situational intentions aligning with their domain expertise.
\end{researchobjective}

\begin{researchobjective}{ro:3}
Facilitate the domain customizability of situational applications, allowing domain experts to address their domain-specific requirements using context aware functionalities. 
\end{researchobjective}


\vspace{-15pt}
\hypertarget{contributions}{%
\section{Contributions}\label{sec:introduction.contributions}}
\vspace{15pt}

The main contribution of this thesis is an EUD-based methodology for domain-specific situational applications. The solution is designed to assist non-technical domain experts in decision-making by enabling the efficient development of web compositions, with a dedicated focus on domain specificity. The thesis yields number of different developments to meet the aforementioned objectives:

\begin{itemize}
\tightlist
\item
  \textbf{Enabled development of useable situational decision-making
    tools by domain experts}. This thesis presents a set of methods and
    mechanisms to assist domain expert in decision-making process by
    facilitating automatic \gls{api} composition and web mashups deployment
    (cf. Chapter 5) . The tool development complies component-based and
    goal-driven development principles and consists of multiple model
    transformations.
    
\item
  \textbf{Enabled conversational interaction mechanisms at the User
  Interface \gls{ui} level.} \gls{disco} \autocite{Zarei2021} \autocite{Zarei2020} is the proposed tool to increase
  expressiveness power by enabling natural language interaction during
  component discovery and solution composition (cf. Chapter 6). DisCo
  enhances the tool's expressiveness and usability and eliminates any
  unnecessary learning overload since domain experts are not required to
  follow any grammatical rules.
  
\item
  \textbf{Enabled domain-specific customization} This thesis introduces
  a mechanism, which early works introduced in \autocite{Zarei2018} \autocite{Zarei2018a}, to facilitate domain model generation
  and domain-specific customization during both development and runtime,
  ensuring an adequate level of domain specificity (cf. Chapter 7). It
  includes tools for generating domain models leveraging \gls{llm} for
  enhancing end result and integrating domain layer across various
  levels such as components and \gls{ui}.
\end{itemize}


\vspace{-15pt}
\hypertarget{sec:scope}{%
\section{Scope}\label{sec:scope}}
\vspace{15pt}

The scope of this thesis encompasses a comprehensive study of domain-specific situational applications, with a focus on empowering domain experts in decision-making processes by addressing objectives \cref{ro:1} to \cref {ro:3}. This work primarily concentrates on the fields of Web Engineering as well as \gls{hci}, \gls{eud}, and Semantic Web. The research concludes in the development of an \gls{eud} solution to facilitate the creation of web mashups tailored to specific domains. The following research challenges are fallen outside the scope of this thesis:

\begin{itemize}
\tightlist
\item
  Since domain specificity is the central focus point of this thesis,
  developing a general-purpose web mashup is not addressed by this work.
  Although domain models and configuration tools proposed here can be
  used either as standalone models or domain layers to adopt
  general-purpose mashups to a specific domain, the adoption provides
  different usage scenario that is not addressed here.
\item
  This thesis focuses on assisting domain experts in situational
  decision-making by simplifying access to online data and
  functionalities and enabling their composition. The scope of this work
  does not include database and document management, business model
  processing, or the provision of expert capabilities.
\item
  Although designing uniform and responsive \gls{ui} for
  components is a crucial factor for enhancing the usability, the
  responsibility lies on component developers. This thesis partially
  tackles this issue by providing component generation and composition
  mechanisms.
\item
  The focus of this thesis is on the state-of-the-art technologies;
  therefore, we consider the \gls{api}s as the building blocks for components
  and composition.
\end{itemize}


\vspace{-10pt}
\hypertarget{structure-of-this-thesis}{%
\section{Thesis Outline}\label{structure-of-this-thesis}}
\vspace{15pt}

The rest of this thesis is structured as follows: Chapter2 provides a comprehensive investigation into the process of identifying the central problem. To underscore the research problem's significance and relevance, a motivation scenario is designed, and key stakeholders affected by the problem are identified. This process guides the requirement elicitation for a  suitable \gls{eud} solution. Chapter3 reviews state-of-the-art research in the field of \gls{eud} with a focus on situational decision-making and domain specificity. It also introduces an assessment framework to evaluate the studied solution categories against requirements introduced in Chapter2. Chapter4 conceptualizes the domain-specific situational EUD application by designing the necessary toolkit and mechanisms to address the central problem fully. Chapters 5 through 7 address the three main research objectives. Chapter5 focuses on usability by introducing a set of methods for generating and composing components and presents the overall solution architecture. Chapter6 proposes the \gls{disco} to enhance the solution's expressiveness. DisCo uses the query expansion technique on the input to assist domain experts during the discovery and composition phases. The methods for domain-specific tool configuration such as domain model generation and implementation are presented in Chapter 7. The overall approach is evaluated in Chapter8 followed by a conclusion and points for future research direction in  Chapter9.


\pagebreak

\vspace{-15pt}
\hypertarget{summary}{%
\section{Summary}\label{summary}}
\vspace{15pt}

This chapter sets the groundwork for this thesis by providing a comprehensive overview of the research context in the field of \gls{eud}, with an emphasize on the domain specificity of situational applications. The central problem and contributing subproblems are identified and the research objectives were aligned accordingly. Three subproblems each affecting the central problem by targeting usability, expressiveness, and domain specificity. The following chapter further explores the central problem by examining a motivation scenario and extracting requirements for domain-specific situational application tool. 